Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-06 formally.tex

The rules above are numerous and fairly uninteresting. That makes them quite hazardous -- the editor designer might make a mistake absent-mindedly. One check against this is to prove a lemma that establishes that movement actions move the cursor but do not change the cursor erasure.\todo{make sure ian proved this }
\begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure} \todo{make lemmas number with theorems} 
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\performMove{\ztau}{\aMove{\delta}}{\ztau'}$ then $\removeSel{\ztau}=\removeSel{\ztau'}$.
\item If $\performMove{\zexp}{\aMove{\delta}}{\zexp'}$ then $\removeSel{\zexp}=\removeSel{\zexp'}$.
\end{enumerate}
\end{lemma}
\noindent Lemma \ref{lemma:movement-erasure} is useful also in that the relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its application.

A more thorough check would be to establish \emph{reachability}, i.e. that it is possible, through a sequence of movement actions, to move the cursor from any position to any other position within an H-expression.\todo{pending ian progress}

@@

\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

Again, the rule for Z-types is self-explanatory:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\htau}
    }{
      \aDel
    }{
      \zwsel{\tehole}
    }
  }
\end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performTyp{\tarr{\ztau}{\htau}}{\aDel}{\tarr{\ztau'}{\htau}}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performTyp{\tarr{\htau}{\ztau}}{\aDel}{\tarr{\htau}{\ztau'}}
%   }
% \end{equation}
\end{subequations}

@@

\paragraph{Deletability}\todo{do we care?}
The meaning of the deletion action is to remove the entire term in focus,
and its definition does not depend on the structure of the term being
deleted. Therefore, there is no theorem to state that acts as a checksum on
the rules analogous to Theorem \ref{thrm:reach} or \ref{thrm:construct},
since we could not have forgotten any rules corresponding to the structure
of the terms.

\subsubsection{Construction}\label{sec:construction} The construction actions, $\aConstruct{\psi}$, are used to construct terms of a shape indicated by $\psi$ into the program at or around the focus.

>>>>

The rules above are numerous and fairly uninteresting. That makes them quite hazardous -- we might make a mistake absent-mindedly. One check against this is to prove a lemma that establishes that movement actions move the cursor but do not change the cursor erasure.
\begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure} \todo{make lemmas number with theorems} 
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\performMove{\ztau}{\aMove{\delta}}{\ztau'}$ then $\removeSel{\ztau}=\removeSel{\ztau'}$.
\item If $\performMove{\zexp}{\aMove{\delta}}{\zexp'}$ then $\removeSel{\zexp}=\removeSel{\zexp'}$.
\end{enumerate}
\end{lemma}
\noindent Lemma \ref{lemma:movement-erasure} is useful also in that the relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its application.

Another check is to establish \emph{reachability}, i.e. that it is possible, through a sequence of movement actions, to move the cursor from any position to any other position within a well-typed H-expression.

This requires developing machinery for sequences of actions. There are two possibilities: we can either add a sequencing action, $\alpha; \alpha$, directly to the syntax of actions, or we can define auxiliary syntax for sequences of actions, $\bar{\alpha}$, and ``multi\-step'' action judgements. To keep the core of the action semantics small, we took the latter approach in Figure \ref{fig:multistep}.

\begin{figure}
$\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\\
\fbox{$\performTyp{\ztau}{\bar{\alpha}}{\ztau'}$}
\begin{mathpar}
\inferrule{ }{
    \performTyp{\ztau}{\cdot}{\ztau}
}

\inferrule{
  \performTyp{\ztau}{\alpha}{\ztau'}\\
  \performTyp{\ztau'}{\bar{\alpha}}{\ztau''}
}{
  \performTyp{\ztau}{\alpha; \bar{\alpha}}{\ztau''}
}
\end{mathpar}
\fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
\begin{mathpar}
\inferrule{ }{
  \performSyn{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}{\htau}
}

\inferrule{
  \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}\\\\
  \performSyn{\hGamma}{\zexp'}{\htau'}{\bar{\alpha}}{\zexp''}{\htau''}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\alpha; \bar{\alpha}}{\zexp''}{\htau''}
}
\end{mathpar}
\fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
\begin{mathpar}
\inferrule{ }{
  \performAna{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}
}

\inferrule{
  \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}\\\\
  \performAna{\hGamma}{\zexp'}{\htau}{\bar\alpha}{\zexp''}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\alpha; \bar\alpha}{\zexp''}
}
\end{mathpar}
\caption{Multistep Action Performance}
\label{fig:multistep}
\end{figure}

With these definitions, we can state reachability as follows:

\begin{theorem}[Reachability] ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\removeSel{\ztau}=\removeSel{\ztau'}$ then there exists some $\bar\alpha$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performTyp{\ztau}{\bar\alpha}{\ztau'}$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$. 
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zexp'}{\htau}$.
\end{enumerate}
\end{theorem}

To prove Reachability, we break it into two lemmas. One...\todo{write this}

@@

\subsubsection{Deletion} The $\aDel$ action replaces the cursor with an empty hole.

The type action rule is self-explanatory:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\htau}
    }{
      \aDel
    }{
      \zwsel{\tehole}
    }
  }
\end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performTyp{\tarr{\ztau}{\htau}}{\aDel}{\tarr{\ztau'}{\htau}}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performTyp{\tarr{\htau}{\ztau}}{\aDel}{\tarr{\htau}{\ztau'}}
%   }
% \end{equation}
\end{subequations}

@@

\subsubsection{Construction}\label{sec:construction} The construction actions, $\aConstruct{\psi}$, are used to construct terms of a shape indicated by $\psi$ into the program at or around the cursor.
